---
title: "HW - Tidyverse consolidation"
output: html_notebook
---

*Question 1*
Load the code_clan_tweets.csv data. Find the number of rows, columns, and list all the variable names.

```{r}
library(tidyverse)

rtweets <- read_csv("data/code_clan_tweets.csv")

rtweets
```
a) Number of rows = 234
b) Number of columns = 27
c) Variable names:
```{r}
rtweets %>%
  ls()
```

*Question 2*
Find the total number of favourites (stored in favorite_count) that CodeClan tweets have got. Don’t add a tweet’s favorite_count to the total when the tweet was quoted (see the is_quote variable).
```{r}
rtweets %>%
  count(favorite_count)
  as.numeric
```
```{r}
275+50+22+30+23+34
```

```{r}
  
rtweets$is_quote [rtweets$is_quote == "TRUE"] <- 1
rtweets$is_quote [rtweets$is_quote == "FALSE"] <- 0
rtweets$is_quote <- as.integer(rtweets$is_quote)

rtweets
```
```{r}
rtweets%>%

# Can't work out how to extract and line up columns "is_favorite" (11)n  & is_quote (13), then multiply values, but this is what I've been trying to do
  
rtweets = subset(cols_only(), select = -c(favorite_count, is_quote))
rtweets = subset(rtweets, select = -c(k, m))
```

```{r}
rtweets%>%
  select(favorite_count)
  dplyr::filter(is_quote = TRUE)
  
```


*Question 3*
Summarise the mean number of retweets (retweet_count), split by the type of platform that was used to tweet (source). Again, omit quoted tweets from the mean.



