---
title: "Tidy Data Mini-Lab"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---


Now it's your turn to practice the verbs on a new dataset. Complete the following task:

```{r}
library(tidyverse)
library(tidyr)
```

**Task 1.**
a) Load in the `country_data.csv` file and take a look at it.
```{r}
country_data <- read_csv("4_tidy_data_data/country_data.csv")
```
b)  What needs changed in this to make it tidy?
The following to do not accord with tidy data structure:
(i) Rule 3 - Each value must have its own cell. The first column violates this rule values are sharing a cell (eg 1, Afghanistan). 
=> The values need to be separated.

(ii) Rule 1 - Each variable must have its own column. This is violated as the variables "population" and "cases" are in the same column ("type"). 
=> This column should be removed, and replaced by two columns, one for "population" and one for "cases".

c)Which columns need made into long format, and which need spread to rows?
#"Country" and "Type" need to be made into long format???
the years need to be made into rows

**Task 2.**\
Do the appropriate pivoting of your columns using R
```{r}
library(tidyr)
pivot_longer(cols = c("Y1999", "Y2000", "Y2001"), names_to = "year", values_to = "year") 
pivot_wider(id_cols = c("population", "cases")) names_from = "type", values_from = "year")
```
--------------------------Corrections-------------------------------------
Do the appropriate pivoting of your columns using R
 
```{r}
library(tidyverse)

# pivot longer expects:
pivot_longer(data =, cols =, names_to = , values_to = )

# so the data is the first argument
pivot_longer(country_data, cols = c("Y1999", "Y2000", "Y2001"), names_to = "year", values_to = "value")
```

# alternatively we can pipe the data in. When piping: the left hand side of the pipe, in this case country_data gets placed into the first argument (data) of the function on the right hand side of the pipe. Because the tidyverse functions often have the data as the first argument, piping is very natural. We can then pipe again into the pivot wider function. This time our data will be the result of our pivot longer...

country_data %>% 
  pivot_longer(cols = c("Y1999", "Y2000", "Y2001"), names_to = "year", values_to = "value") %>% 
  pivot_wider(names_from = "type", values_from = "value")

# This is equivalent to

pivot_wider(
  data = pivot_longer(
    data = country_data,
    cols = c("Y1999", "Y2000", "Y2001"),
    names_to = "year",
    values_to = "value"
    ),
  names_from = "type",
  values_from = "value"
)


**Task 3.**\
Separate the `country` variable so you have the name in one column and the number in another.
```{r}
separate("country", c("number", "name"), sep = ".")
```
-------------------------------------------------------------------------------
Task 3
Separate the country variable so you have the name in one column and the number in another.
{r}
# again, the data is the first argument:
separate(data = country_data, "country", c("number", "name"), sep = "\\.")

# equivalent to
country_data %>% 
  separate("country", c("number", "name"), sep = "\\.")



**Task 4.**\
Unite the `code_letter` and `code_numb` variables so that they are both in one column, with no spaces separating them.
```{r}
unite(code, col = c("code_letter", "code_numb"), sep = 
```

-------------------------------
```{r}
# again, the data is the first argument
unite(data = country_data, col = "code", c("code_letter", "code_numb"), sep = '')

```



# equivalent to

country_data %>% 
  unite(col = "code", c("code_letter", "code_numb"), sep = '')